---
title: "HR Analytics Report"
author: "L. di Bernardo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# Set global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## R Markdown

This report provides an analysis of HR analytics data, exploring key metrics such as employee demographics, job satisfaction, salary distribution, and turnover intentions. 
The codes under the hood have been left visible on purpose to enable a view of the work underneath.

```{r}
# Load necessary libraries
library(tidyverse)

# Load the dataset
hr_data <- read.csv("hr_analytics_mockup.csv")

# Preview the dataset

# Summary statistics
summary(hr_data)

# Check for missing values
colSums(is.na(hr_data))
```

## Employee Demogrpahics

# Gender Distribution

```{r}

# Gender distribution
gender_counts <- hr_data %>% count(Gender)

# Plot gender distribution
gender_counts %>%
  ggplot(aes(x = Gender, y = n, fill = Gender)) +
  geom_bar(stat = "identity") +
  labs(title = "Gender Distribution", x = "Gender", y = "Count") +
  theme_minimal()
```

# Department Breakdown

```{r}
# Department distribution
department_counts <- hr_data %>% count(Department)

# Plot department distribution
department_counts %>%
  ggplot(aes(x = reorder(Department, -n), y = n, fill = Department)) +
  geom_bar(stat = "identity") +
  labs(title = "Employees by Department", x = "Department", y = "Count") +
  theme_minimal()

```

## Salary Analysis

# Salary Distrubution

```{r}
# Salary distribution plot
hr_data %>%
  ggplot(aes(x = MonthlySalary)) +
  geom_histogram(binwidth = 500, fill = "skyblue", color = "black") +
  labs(title = "Salary Distribution", x = "Monthly Salary", y = "Frequency") +
  theme_minimal()

```

# Salary by Department

```{r}
# Boxplot of salary by department
hr_data %>%
  ggplot(aes(x = Department, y = MonthlySalary, fill = Department)) +
  geom_boxplot() +
  labs(title = "Salary by Department", x = "Department", y = "Monthly Salary") +
  theme_minimal()

```

## Turnover Analysis

# Turnover Rate 

```{r}
# Turnover rate calculation
turnover_rate <- mean(hr_data$TurnoverIntention) * 100
paste("Turnover Rate: ", round(turnover_rate, 2), "%")

```

# Turnover by Job Satisfaction
```{r}
# Turnover by job satisfaction
hr_data %>%
  ggplot(aes(x = as.factor(JobSatisfaction), fill = as.factor(TurnoverIntention))) +
  geom_bar(position = "fill") +
  labs(title = "Turnover by Job Satisfaction", x = "Job Satisfaction (1-5)", y = "Proportion", fill = "Turnover Intention") +
  theme_minimal()

```

# Insight and Recommendations

1. Demographics: Gender distribution and department representation highlight diversity in the workforce.
2. Salary Analysis: Salary disparities between departments may need attention to ensure equity.
3. Turnover Analysis: Job satisfaction is strongly associated with turnover intention; initiatives to improve satisfaction may reduce attrition.

# Conclusion

This analysis provides actionable insights into workforce dynamics. Future steps could include predictive modeling to identify employees at risk of leaving and targeted initiatives to improve employee retention.


